---
title: "Traveller profile clustering: Gower + PAM"
format: html
---

```{r setup}
#| message: false
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cluster)    # daisy (Gower) and pam
library(factoextra) # cluster visualisation
```

```{r load}
source(here::here("proj/munge/01_cluster_prep.R"))
```

## Prepare clustering variables

`modal_choice` is passed as a factor so Gower distance treats it as categorical. All continuous variables are left on their natural scale — Gower normalises each variable to [0,1] internally before combining distances, so manual standardisation is not needed.

```{r prep}
cluster_input <- indiv_cluster_df |>
  select(
    IndividualID,
    # Volume & distance
    n_trips, total_distance, share_long_trips,
    # Modal choice (categorical)
    modal_choice,
    # Mode shares
    share_walk, share_cycle, share_car_driver, share_car_pass,
    share_bus, share_rail, share_other,
    # Purpose mix
    share_work, share_education, share_escort_educ,
    share_shopping, share_escort, share_leisure
  ) |>
  mutate(modal_choice = factor(modal_choice, levels = 1:7,
                                labels = c("Walk", "Cycle", "Car driver",
                                           "Car passenger", "Bus", "Rail", "Other"))) |>
  drop_na()

cat("Individuals after dropping NA:", nrow(cluster_input), "\n")
```

## Subsample for distance matrix computation

Gower distance requires an n×n matrix which is memory-prohibitive for large samples. We compute Gower + PAM on a subsample, then assign the full sample to the nearest medoid.

```{r subsample}
set.seed(42)
sample_size <- 5000

sample_idx <- sample(nrow(cluster_input), min(sample_size, nrow(cluster_input)))
cluster_sample <- cluster_input[sample_idx, ]

cat("Sample size:", nrow(cluster_sample), "\n")
cat("Full dataset size:", nrow(cluster_input), "\n")
```

## Compute Gower distance matrix on subsample

```{r gower}
gower_dist <- daisy(
  cluster_sample |> select(-IndividualID),
  metric = "gower"
)

summary(gower_dist)
```

## Determine number of clusters

Silhouette width across k = 2 to 8 on the subsample:

```{r silhouette}
#| fig-width: 7
#| fig-height: 4
sil_width <- numeric(8)

for (k in 2:8) {
  pam_fit_k <- pam(gower_dist, k = k, diss = TRUE)
  sil_width[k] <- pam_fit_k$silinfo$avg.width
}

tibble(k = 2:8, silhouette = sil_width[2:8]) |>
  ggplot(aes(x = k, y = silhouette)) +
  geom_line() +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2:8) +
  labs(x = "Number of clusters (k)", y = "Average silhouette width") +
  theme_minimal()
```

Optimal k based on silhouette:

```{r optimal-k}
optimal_k <- which.max(sil_width[2:8]) + 1
cat("Optimal k:", optimal_k, "\n")
```

## Fit PAM with optimal k on subsample

```{r pam}
pam_fit <- pam(gower_dist, k = optimal_k, diss = TRUE)
```

## Assign full sample to nearest medoid

Extract medoid profiles from the subsample, then for each individual in the full dataset compute Gower distance to each medoid and assign to the closest one.

```{r assign-full}
# Extract medoid rows from the subsample
medoid_profiles <- cluster_sample[pam_fit$id.med, ] |>
  select(-IndividualID)

# For each individual in the full dataset, compute distance to each medoid
# and assign to the nearest one
cluster_vars <- cluster_input |> select(-IndividualID)

assign_to_medoid <- function(data, medoids) {
  sapply(seq_len(nrow(data)), function(i) {
    dists <- sapply(seq_len(nrow(medoids)), function(m) {
      daisy(rbind(data[i, ], medoids[m, ]), metric = "gower")[1]
    })
    which.min(dists)
  })
}

indiv_clustered <- cluster_input |>
  mutate(cluster = factor(assign_to_medoid(cluster_vars, medoid_profiles)))

cat("Cluster sizes (full sample):\n")
print(table(indiv_clustered$cluster))
```

## Cluster summaries

Medoid profiles — each row is a real individual representing their cluster:

```{r medoids}
cluster_sample[pam_fit$id.med, ] |>
  mutate(cluster = 1:optimal_k) |>
  select(cluster, everything(), -IndividualID)
```

Variable means by cluster (full sample):

```{r cluster-means}
indiv_clustered |>
  group_by(cluster) |>
  summarise(
    n = n(),
    across(c(n_trips, total_distance, share_long_trips,
             share_walk, share_cycle, share_car_driver, share_car_pass,
             share_bus, share_rail, share_other,
             share_work, share_education, share_escort_educ,
             share_shopping, share_escort, share_leisure),
           mean, na.rm = TRUE)
  )
```

Modal choice distribution by cluster (full sample):

```{r modal-by-cluster}
indiv_clustered |>
  count(cluster, modal_choice) |>
  group_by(cluster) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = modal_choice, y = pct, fill = cluster)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "% within cluster", fill = "Cluster") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Silhouette plot

Based on the subsample:

```{r silhouette-plot}
#| fig-width: 7
#| fig-height: 6
fviz_silhouette(pam_fit)
```

## Stability check

Re-run on 5 different subsamples of the same size and compare cluster sizes:

```{r stability}
set.seed(42)
seeds <- sample(1000, 5)

stability <- lapply(seeds, function(s) {
  set.seed(s)
  idx <- sample(nrow(cluster_input), sample_size)
  samp <- cluster_input[idx, ]
  dist_s <- daisy(samp |> select(-IndividualID), metric = "gower")
  fit_s <- pam(dist_s, k = optimal_k, diss = TRUE)
  as.data.frame(t(table(fit_s$clustering))) |>
    select(cluster = Var2, n = Freq)
}) |>
  bind_rows(.id = "run")

stability
```

---

## Subclustering within macro-clusters

For each macro-cluster, repeat Gower + PAM independently to find subclusters. Silhouette chooses k up to a maximum of 5 per macro-cluster.

```{r subcluster-fn}
# k_override: named list to force k for specific macro-clusters,
#   e.g. list("1" = 4). NULL means use silhouette.
# flat_tol: if max - min silhouette across k is below this threshold,
#   the cluster is considered homogeneous and left unsplit.
subcluster_fn <- function(data, macro_id, sample_size = 5000, k_max = 5,
                           k_override = NULL, flat_tol = 0.01, seed = 42) {
  set.seed(seed)
  cat("\n--- Macro-cluster", macro_id, "| n =", nrow(data), "---\n")

  cluster_vars <- data |> select(-IndividualID)

  # Subsample if needed
  if (nrow(data) > sample_size) {
    idx <- sample(nrow(data), sample_size)
    samp <- cluster_vars[idx, ]
  } else {
    samp <- cluster_vars
  }

  # Gower distance on sample
  dist_s <- daisy(samp, metric = "gower")

  # Silhouette across k = 2:k_max
  sil <- numeric(k_max)
  for (k in 2:k_max) {
    fit_k <- pam(dist_s, k = k, diss = TRUE)
    sil[k] <- fit_k$silinfo$avg.width
  }

  # Plot silhouette
  print(
    tibble(k = 2:k_max, silhouette = sil[2:k_max]) |>
      ggplot(aes(x = k, y = silhouette)) +
      geom_line() + geom_point(size = 3) +
      scale_x_continuous(breaks = 2:k_max) +
      labs(title = paste("Macro-cluster", macro_id),
           x = "k", y = "Avg silhouette width") +
      theme_minimal()
  )

  # Check for flat silhouette — skip subclustering if homogeneous
  sil_range <- max(sil[2:k_max]) - min(sil[2:k_max])
  if (sil_range < flat_tol & is.null(k_override[[as.character(macro_id)]])) {
    cat("Flat silhouette (range =", round(sil_range, 4),
        ") — leaving cluster unsplit.\n")
    return(data |> mutate(subcluster = paste0(macro_id, ".1")))
  }

  # Determine k: manual override takes precedence over silhouette
  if (!is.null(k_override[[as.character(macro_id)]])) {
    optimal_k <- k_override[[as.character(macro_id)]]
    cat("Using manual override k =", optimal_k, "\n")
  } else {
    optimal_k <- which.max(sil[2:k_max]) + 1
    cat("Optimal k (silhouette):", optimal_k, "\n")
  }

  # Fit PAM with chosen k
  pam_sub <- pam(dist_s, k = optimal_k, diss = TRUE)

  if (nrow(data) > sample_size) {
    medoids    <- samp[pam_sub$id.med, ]
    sub_labels <- assign_to_medoid(cluster_vars, medoids)
  } else {
    sub_labels <- pam_sub$clustering
  }

  data |>
    mutate(subcluster = paste0(macro_id, ".", sub_labels))
}
```

```{r subclustering}
#| message: false
# Pass k_override as a named list: cluster label (as string) -> forced k
k_overrides <- list("1" = 4)

subclustered <- indiv_clustered |>
  group_by(cluster) |>
  group_modify(~ subcluster_fn(.x, macro_id = .y$cluster,
                                k_override = k_overrides)) |>
  ungroup()
```

## Subcluster summaries

Sizes of all final subclusters:

```{r subcluster-sizes}
subclustered |>
  count(cluster, subcluster) |>
  arrange(cluster, subcluster)
```

Variable means by subcluster:

```{r subcluster-means}
subclustered |>
  group_by(subcluster) |>
  summarise(
    n = n(),
    across(c(n_trips, total_distance, share_long_trips,
             share_walk, share_cycle, share_car_driver, share_car_pass,
             share_bus, share_rail, share_other,
             share_work, share_education, share_escort_educ,
             share_shopping, share_escort, share_leisure),
           mean, na.rm = TRUE)
  ) |>
  arrange(subcluster)
```

Modal choice distribution by subcluster:

```{r subcluster-modal}
#| fig-width: 9
#| fig-height: 5
subclustered |>
  count(subcluster, modal_choice) |>
  group_by(subcluster) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = modal_choice, y = pct, fill = subcluster)) +
  geom_col(position = "dodge") +
  facet_wrap(~ subcluster, scales = "free_y") +
  labs(x = NULL, y = "% within subcluster") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```
